---
title: 'Fundraising Analytics on DonorsChoose.org data'
author: "Alejandra Gerosa"
date: "12/7/2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(lubridate)
library(scales)
library(cowplot)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## The project

This project analyses [DonorsChoose.org](www.donorschoose.org)'s donation data as released in their mid-2018 [Kaggle competition](https://www.kaggle.com/donorschoose/io). For this writeup, I set out to summarize a few key insights regarding their fundrasing performance between 2013 and 2017. For a more detailed approach, you can dig into the full code of my ongoing exploration of the dataset [here](https://github.com/alegerosa/DonorsChoose.org). 

```{r include=FALSE}
df_paths <- list.files("data", full.names = TRUE)
list_of_dfs <- lapply(df_paths, read_csv)
df_names <- tolower(basename(substr(df_paths, 1, nchar(df_paths)-4)))
names(list_of_dfs) <- df_names
for (i in 1:length(list_of_dfs)) {
  list_of_dfs[[i]] <- rename_all(list_of_dfs[[i]], funs(gsub(" ", "_", .)))
  list_of_dfs[[i]] <- rename_all(list_of_dfs[[i]], funs(tolower(.)))
} 
list2env(list_of_dfs, .GlobalEnv)
donations$donation_included_optional_donation <- donations$donation_included_optional_donation == "Yes"
donors$donor_is_teacher <- donors$donor_is_teacher == "Yes"
projects$project_type <- as.factor(projects$project_type)
projects$project_grade_level_category <- as.factor(projects$project_grade_level_category)
projects$project_resource_category <- as.factor(projects$project_resource_category)
projects$project_current_status <- as.factor(projects$project_current_status)
schools$school_metro_type <- as.factor(schools$school_metro_type)
teachers$teacher_prefix <- as.factor(teachers$teacher_prefix)
donations_duplicates <- donations %>%
  filter(duplicated(donation_id) | duplicated(donation_id, fromLast = TRUE)) %>%
  arrange(donation_id)
donations <- donations %>% filter(!duplicated(donation_id))
by_project <- donations %>% group_by(project_id) %>%
  summarize(count_donations = n(),
            sum_donations = sum(donation_amount),
            avg_amount = mean(donation_amount)) %>%
  arrange(desc(count_donations)) 
by_donor_id <- donations %>% group_by(donor_id) %>%
  summarize(count_donations = n(),
            sum_donations = sum(donation_amount),
            avg_amount = mean(donation_amount)) %>%
  arrange(desc(count_donations)) 

ids_of_onetime_donors <- select(filter(by_donor_id, count_donations == 1), donor_id)
ids_of_repeat_donors <- select(filter(by_donor_id, count_donations > 1), donor_id)

donations <- donations %>%
  mutate(
    first_donation = case_when(
      donor_cart_sequence == 1 ~ "first",
      donor_cart_sequence > 1 ~ "recurrent"),
    one_and_done = case_when(
      as.character(donor_id) %in% pull(ids_of_onetime_donors) ~ "One-and-done donor",
      as.character(donor_id) %in% pull(ids_of_repeat_donors) ~ "Recurrent donor")
    )
donations_plus <- donations %>%
  left_join(projects, by = "project_id") %>%
  left_join(donors, by = "donor_id") %>%
  left_join(schools, by = "school_id") %>%
  left_join(teachers, by = "teacher_id")

revenue_per_year_plot <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  ggplot(aes(x = year(donation_received_date), y = donation_amount/1000000, fill = "pink")) +
  geom_col() +
  labs(y = "Total donation value (in millions)") +
  theme(legend.position = "none",
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = dollar_format())
yearly_growth_table <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  group_by(year = year(donation_received_date)) %>%
  summarize(total = sum(donation_amount)) %>%
  mutate(perc_growth = (total/lag(total))-1)
cagr <- percent(mean(yearly_growth_table$perc_growth, na.rm = TRUE))
perc_growth_per_year_plot <- yearly_growth_table %>%
  ggplot(aes(x = year, y = perc_growth)) +
  geom_line() +
  labs(y = "% growth") +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = c(0, .25, .5)) +
  coord_cartesian(ylim = c(0,.5)) +
  geom_text(aes(label = percent(perc_growth)), position = position_dodge(), vjust = -1.5)
perc_growth_per_year_plot

donors_before_data <- donations_plus %>%
  group_by(donor_id) %>%
  filter(min(donor_cart_sequence) > 1) %>%
  select(donor_id)
donors_2012 <- donations_plus %>%
  filter(year(donation_received_date) == 2012) %>%
  select(donor_id)
donors_2013 <- donations_plus %>%
  filter(year(donation_received_date) == 2013) %>%
  select(donor_id)
donors_2014 <- donations_plus %>%
  filter(year(donation_received_date) == 2014) %>%
  select(donor_id)
donors_2015 <- donations_plus %>%
  filter(year(donation_received_date) == 2015) %>%
  select(donor_id)
donors_2016 <- donations_plus %>%
  filter(year(donation_received_date) == 2016) %>%
  select(donor_id)
donors_2017 <- donations_plus %>%
  filter(year(donation_received_date) == 2017) %>%
  select(donor_id)
donors_2018 <- donations_plus %>%
  filter(year(donation_received_date) == 2018) %>%
  select(donor_id)


donations_plus <- donations_plus %>%
  mutate(donor_gave_last_year = case_when(
    year(donation_received_date) == 2013 ~ as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2014 ~ as.character(donor_id) %in% pull(donors_2013),
    year(donation_received_date) == 2015 ~ as.character(donor_id) %in% pull(donors_2014),
    year(donation_received_date) == 2016 ~ as.character(donor_id) %in% pull(donors_2015),
    year(donation_received_date) == 2017 ~ as.character(donor_id) %in% pull(donors_2016),
    year(donation_received_date) == 2018 ~ as.character(donor_id) %in% pull(donors_2017)))

donations_plus <- donations_plus %>%
  mutate(donor_gave_ever_before = case_when(
    year(donation_received_date) == 2012 ~ as.character(donor_id) %in% pull(donors_before_data),
    year(donation_received_date) == 2013 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2014 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2013) | as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2015 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2014) | as.character(donor_id) %in% pull(donors_2013) | as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2016 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2015) | as.character(donor_id) %in% pull(donors_2014) | as.character(donor_id) %in% pull(donors_2013) | as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2017 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2016) | as.character(donor_id) %in% pull(donors_2015) | as.character(donor_id) %in% pull(donors_2014) | as.character(donor_id) %in% pull(donors_2013) | as.character(donor_id) %in% pull(donors_2012),
    year(donation_received_date) == 2018 ~ as.character(donor_id) %in% pull(donors_before_data) | as.character(donor_id) %in% pull(donors_2017) | as.character(donor_id) %in% pull(donors_2016) | as.character(donor_id) %in% pull(donors_2015) | as.character(donor_id) %in% pull(donors_2014) | as.character(donor_id) %in% pull(donors_2013) | as.character(donor_id) %in% pull(donors_2012)))

donations_plus <- donations_plus %>%
  mutate(retention_status = case_when(
    donor_gave_last_year ~ "Retained donor",
    !donor_gave_last_year & donor_gave_ever_before ~ "Reactivated donor",
    !donor_gave_ever_before ~ "New donor"
  ))

revenue_per_year_per_retained <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  ggplot(aes(x = year(donation_received_date), y = donation_amount/1000000, fill = retention_status)) +
  geom_col() +
  labs(y = "Total donation value (in millions)", fill = "Donor gave in the year prior?") +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank()) +
  scale_y_continuous(labels = dollar_format())
yearly_proper_retained_table <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  group_by(year = year(donation_received_date), donor_gave_last_year) %>%
  summarize(total = sum(donation_amount)) %>%
  spread(donor_gave_last_year, total) %>%
  mutate(total = sum(c(`FALSE`,`TRUE`), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(perc_retained = `TRUE`/lag(total))
avg_retention <- percent(mean(yearly_proper_retained_table$perc_retained, na.rm = TRUE))
perc_proper_retained_per_year_plot <- yearly_proper_retained_table %>%
  ggplot(aes(x = year, y = perc_retained)) +
  geom_line() +
  labs(y = "Retention Rate") +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = c(.3, .6)) +
  coord_cartesian(ylim = c(.3,.6)) +
  geom_text(aes(label = percent(perc_retained)), position = position_dodge(), vjust = -1.5)

```

## Key Insights
* Donations are growing by an average of `r cagr` every year, and most donations come from recurring donors.
* Top 10 states account for 56% of donations, and top 3 states account for ~30%.
* December, August and September account for ~35% of donations.
* Most donors have given only once.

## On average, donations are growing by `r cagr` each year

```{r}
plot_grid(perc_growth_per_year_plot, revenue_per_year_plot, align = "v", nrow = 2, rel_heights = c(1/4, 3/4))
            

```

The good news is that the total amount raised for teacher's projects is growing every year at a rate of almost 25%. The somewhat less positive news is that growth is slowing down, with only 16% in 2017. How worrisome this is, if at all, depends on the expectation and intention of the organization. Was this a surprise or is it an expected tapering off, a stabilization following a soar? What was the growth goal? Were there growth-oriented actions that performed below earlier years' results? If so, can we pinpoint the reasons?. In any case, if the goal is to maintain 25% growth, however, further action will be needed.

Personally, when I look at the health of an organzation's fundraising base there's a question that I care about much more than how fast we are growing. That is: Are our donors coming back?


```{r}
plot_grid(perc_proper_retained_per_year_plot, revenue_per_year_per_retained, align = "v", nrow = 2, rel_heights = c(1/4, 3/4))
```
