---
title: 'Fundraising Analytics on DonorsChoose.org data'
author: "Alejandra Gerosa"
date: "12/7/2018"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(lubridate)
library(scales)
library(cowplot)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## The project

This project analyses [DonorsChoose.org](www.donorschoose.org)'s donation data as released in their mid-2018 [Kaggle competition](https://www.kaggle.com/donorschoose/io). For this writeup, I set out to summarize a few key insights regarding their fundrasing performance between 2013 and 2017. For a more detailed approach, you can dig into the full code of my ongoing exploration of the dataset [here](https://github.com/alegerosa/DonorsChoose.org). 

```{r include=FALSE}
df_paths <- list.files("data", full.names = TRUE)
list_of_dfs <- lapply(df_paths, read_csv)
df_names <- tolower(basename(substr(df_paths, 1, nchar(df_paths)-4)))
names(list_of_dfs) <- df_names
for (i in 1:length(list_of_dfs)) {
  list_of_dfs[[i]] <- rename_all(list_of_dfs[[i]], funs(gsub(" ", "_", .)))
  list_of_dfs[[i]] <- rename_all(list_of_dfs[[i]], funs(tolower(.)))
} 
list2env(list_of_dfs, .GlobalEnv)
donations$donation_included_optional_donation <- donations$donation_included_optional_donation == "Yes"
donors$donor_is_teacher <- donors$donor_is_teacher == "Yes"
projects$project_type <- as.factor(projects$project_type)
projects$project_grade_level_category <- as.factor(projects$project_grade_level_category)
projects$project_resource_category <- as.factor(projects$project_resource_category)
projects$project_current_status <- as.factor(projects$project_current_status)
schools$school_metro_type <- as.factor(schools$school_metro_type)
teachers$teacher_prefix <- as.factor(teachers$teacher_prefix)
donations_duplicates <- donations %>%
  filter(duplicated(donation_id) | duplicated(donation_id, fromLast = TRUE)) %>%
  arrange(donation_id)
donations <- donations %>% filter(!duplicated(donation_id))
by_project <- donations %>% group_by(project_id) %>%
  summarize(count_donations = n(),
            sum_donations = sum(donation_amount),
            avg_amount = mean(donation_amount)) %>%
  arrange(desc(count_donations)) 
by_donor_id <- donations %>% group_by(donor_id) %>%
  summarize(count_donations = n(),
            sum_donations = sum(donation_amount),
            avg_amount = mean(donation_amount)) %>%
  arrange(desc(count_donations)) 

ids_of_onetime_donors <- select(filter(by_donor_id, count_donations == 1), donor_id)
ids_of_repeat_donors <- select(filter(by_donor_id, count_donations > 1), donor_id)

donations <- donations %>%
  mutate(
    first_donation = case_when(
      donor_cart_sequence == 1 ~ "first",
      donor_cart_sequence > 1 ~ "recurrent"),
    one_and_done = case_when(
      as.character(donor_id) %in% pull(ids_of_onetime_donors) ~ "One-and-done donor",
      as.character(donor_id) %in% pull(ids_of_repeat_donors) ~ "Recurrent donor")
    )
donations_plus <- donations %>%
  left_join(projects, by = "project_id") %>%
  left_join(donors, by = "donor_id") %>%
  left_join(schools, by = "school_id") %>%
  left_join(teachers, by = "teacher_id")

revenue_per_year_plot <- donations %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  ggplot(aes(x = year(donation_received_date), y = donation_amount/1000000, fill = "pink")) +
  geom_bar(stat = "sum") +
  labs(y = "Total donation value (in millions)") +
  theme(legend.position = "none",
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = dollar_format())
yearly_growth_table <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  group_by(year = year(donation_received_date)) %>%
  summarize(total = sum(donation_amount)) %>%
  mutate(perc_growth = (total/lag(total))-1)
cagr <- percent(mean(yearly_growth_table$perc_growth, na.rm = TRUE))
perc_growth_per_year_plot <- yearly_growth_table %>%
  ggplot(aes(x = year, y = perc_growth)) +
  geom_line() +
  labs(y = "% growth") +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(ylim = c(0,1))

revenue_per_year_per_first_plot <- donations %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  ggplot(aes(x = year(donation_received_date), y = donation_amount/1000000, fill = first_donation)) +
  geom_bar(stat = "sum") +
  labs(y = "Total donation value (in millions)") +
  theme(legend.position = "bottom",
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = dollar_format())
yearly_retained_table <- donations_plus %>%
  filter(year(donation_received_date) > 2012 & year(donation_received_date) < 2018) %>%
  group_by(year = year(donation_received_date), first_donation) %>%
  summarize(total = sum(donation_amount)) %>%
  spread(first_donation, total) %>%
  mutate(total = first + recurrent) %>%
  ungroup() %>%
  mutate(perc_retained = recurrent/lag(total))
avg_retention <- percent(mean(yearly_retained_table$perc_retained, na.rm = TRUE))
perc_retained_per_year_plot <- yearly_retained_table %>%
  ggplot(aes(x = year, y = perc_retained)) +
  geom_line() +
  labs(y = "% retained") +
  theme(axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(ylim = c(0, 1))
```

## Key Insights
* Donations are growing by an average of `r cagr` every year, and most donations come from recurring donors.
* Top 10 states account for 56% of donations, and top 3 states account for ~30%.
* December, August and September account for ~35% of donations
* Most donors have given only once.
* A lot of projects don't get funded

## On average, donations are growing by `r cagr` each year

```{r}
plot_grid(perc_growth_per_year_plot, revenue_per_year_plot, align = "v", nrow = 2, rel_heights = c(1/4, 3/4))
            

```



```{r}
plot_grid(perc_retained_per_year_plot, revenue_per_year_per_first_plot, align = "v", nrow = 2, rel_heights = c(1/4, 3/4))
```
